\pagestyle{plain}

% \addcontentsline{toc}{section}{Acknowledgements}
{\Large Acknowledgments}\\
\hrule

Any project of this scale could not be completed without the contributions of many individuals. I would like to use this space to acknowledge those who sacrificed much for me to be able to pursue my passion for electronics and who assisted in the formation of this project.

Firstly, I would like to thank my project supervisor, Dr. Winberg, for being a supportive mentor in this endeavour. His enthusiasm and expertise have been a resource that has been invaluable to the creation of PEEPS. 

Secondly, I would like to thank my peers in the Electrical Engineering Department for fostering my passion for engineering, and for the support I have always been given whenever I have needed it. 

Thirdly, I would like to thank the University of Cape Town and the Electrical Engineering Department faculty for providing the resources I have needed to pursue a life surrounded by innovation, science, and creation. Their tutelage and support over the last four years will not be forgotten.

Lastly, but most importantly, I would like to thank my family. Without their dedication, love and sacrifice, I would never have been able to achieve what I have been able to. I am privileged to be in environment where I am encouraged to take risks and explore my interests, while also being supported regardless of whether I succeed or fail. I would like to thank James and Emily in particular for keeping my spirits high no matter the circumstances.
